---
title: "DataAnalysis_HTproject"
author: "Somraj Shrestha"
date: "2026-04-23"
output:
   md_document:
---


===========================================================
OrthoFinder Downstream Analysis
Heat Tolerant vs Non-Heat Tolerant Genomes
============================================================

#Install required packages
```{r}
#install.packages(c("tidyverse", "ggplot2", "ggvenn", "pheatmap", "RColorBrewer", "UpSetR"))
```

#Load the packages
```{r}
library(UpSetR)
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
```

#Load the data
```{r}
ortho <- read.table("Data/Orthogroups.GeneCount.tsv",
                    header = TRUE, sep = "\t", row.names = 1)
ortho <- ortho[, !colnames(ortho) %in% "Total"] #remove Total column if present
meta <- read.csv("Data/metadata.csv")
rownames(meta) <- meta$genome
```


```{r}
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# Confirm column names match metadata
all(colnames(ortho) %in% meta$genome)  # Should return TRUE

# Reorder metadata to match ortho column order
meta <- meta[colnames(ortho), ]

# Separate species by group
heat_species     <- meta$genome[meta$group == "heat_tolerant"]
non_heat_species <- meta$genome[meta$group == "non_heat_tolerant"]
```


#Presence/absence matrix
```{r}
# Convert gene counts to presence (1) / absence (0)
pa_matrix <- as.data.frame((ortho > 0) * 1)

# ============================================================
# 3. CORE, SHARED, AND UNIQUE ORTHOGROUPS
# ============================================================

# --- Core orthogroups: present in ALL 12 genomes ---
core_OGs <- rownames(pa_matrix)[rowSums(pa_matrix) == ncol(pa_matrix)]
cat("Core orthogroups (all 12 genomes):", length(core_OGs), "\n")

# --- Shared between ALL heat tolerant only (present in all 4 HT, absent in all 8 NHT) ---
heat_present     <- rowSums(pa_matrix[, heat_species]) == length(heat_species)
non_heat_absent  <- rowSums(pa_matrix[, non_heat_species]) == 0
heat_unique_strict <- rownames(pa_matrix)[heat_present & non_heat_absent]
cat("Orthogroups UNIQUE to ALL heat tolerant (strict):", length(heat_unique_strict), "\n")

# --- Present in ALL heat tolerant (may also be in some non-HT) ---
heat_core_OGs <- rownames(pa_matrix)[rowSums(pa_matrix[, heat_species]) == length(heat_species)]
cat("Orthogroups in ALL heat tolerant genomes:", length(heat_core_OGs), "\n")

# --- Present in ALL non-heat tolerant only ---
non_heat_present <- rowSums(pa_matrix[, non_heat_species]) == length(non_heat_species)
heat_absent      <- rowSums(pa_matrix[, heat_species]) == 0
non_heat_unique  <- rownames(pa_matrix)[non_heat_present & heat_absent]
cat("Orthogroups UNIQUE to ALL non-heat tolerant:", length(non_heat_unique), "\n")

# --- Softcore: present in at least 3 of 4 heat tolerant ---
heat_softcore <- rownames(pa_matrix)[rowSums(pa_matrix[, heat_species]) >= 3 &
                                     rowSums(pa_matrix[, non_heat_species]) == 0]
cat("Orthogroups in ≥3 heat tolerant, absent in NHT:", length(heat_softcore), "\n")
```

#Save the results
```{r}
# Save results
write.csv(data.frame(Orthogroup = core_OGs),
          "core_orthogroups_all12.csv", row.names = FALSE)
write.csv(data.frame(Orthogroup = heat_unique_strict),
          "orthogroups_unique_heat_tolerant.csv", row.names = FALSE)
write.csv(data.frame(Orthogroup = non_heat_unique),
          "orthogroups_unique_non_heat_tolerant.csv", row.names = FALSE)
write.csv(data.frame(Orthogroup = heat_softcore),
          "orthogroups_softcore_heat_tolerant.csv", row.names = FALSE)
```


#Generating the Upset plot
```{r}
# Presence/absence per species (each species as a set)
upset_data <- as.data.frame(t(pa_matrix))

# Make each species a column (UpSetR needs species as columns)
upset_input <- as.data.frame(pa_matrix)

upsetplot <- upset(upset_input,
      sets = colnames(upset_input),
      nintersects = 25,
      order.by = "freq",
      decreasing = TRUE,
      mb.ratio = c(0.6, 0.4),
      number.angles = 0,
      text.scale = c(1.2, 1.2, 1, 1, 1.2, 1),
      mainbar.y.label = "Number of Orthogroups",
      sets.x.label = "Orthogroups per Species")

upsetplot
```


#HEATMAP — Gene Copy Number Across Species
```{r}

# Use only orthogroups that differ between heat and non-heat
# Filter for variable orthogroups (not all same value)
variable_OGs <- ortho[apply(ortho, 1, var) > 0, ]

# Take top 30 most variable for cleaner heatmap
top_variable <- variable_OGs[order(apply(variable_OGs, 1, var),
                                   decreasing = TRUE), ][1:30, ]

# Annotation for columns (species)
col_annotation <- data.frame(Group = meta$group)
rownames(col_annotation) <- meta$genome

ann_colors <- list(Group = c(heat_tolerant     = "#D55E00",
                              non_heat_tolerant = "#56B4E9"))

heatmap <- pheatmap(top_variable,
         annotation_col = col_annotation,
         annotation_colors = ann_colors,
         color = colorRampPalette(c("white", "#F0E442", "#E74C3C"))(30),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_col = 9,
         fontsize_row = 7,
         main = "Top 30 Variable Orthogroups — Gene Copy Number",
         border_color = NA)
heatmap
```

#Summary Bar Chart
```{r}
summary_df <- data.frame(
  Category = c("Core (all 12)",
               "Heat Tolerant Only (strict)",
               "Non-Heat Tolerant Only",
               "Heat Tolerant Softcore (≥3/4)",
               "Shared (all HT, some NHT)"),
  Count = c(length(core_OGs),
            length(heat_unique_strict),
            length(non_heat_unique),
            length(heat_softcore),
            sum(heat_present) - length(heat_unique_strict))
)


ggplot(summary_df, aes(x = reorder(Category, -Count), y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 0.6) +
  geom_text(aes(label = Count), vjust = -0.5, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00")) +
  labs(title = "Orthogroup Distribution by Category",
       x = NULL, y = "Number of Orthogroups") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
        plot.title = element_text(hjust = 0.5, face = "bold")) 
  
```
- 58 unique orthogroups were found strictly on the heat tolerant species. 

#Future work
The genes within these group should be functionally annotated and any genes that can have potential function relating to heat tolerance can be further studied.

```{r}
print_tree <- function(path = ".", prefix = "") {
  files <- list.files(path, all.files = FALSE)
  
  for (i in seq_along(files)) {
    cat(prefix, "├── ", files[i], "\n", sep = "")
    
    full_path <- file.path(path, files[i])
    
    if (dir.exists(full_path)) {
      print_tree(full_path, paste0(prefix, "│   "))
    }
  }
}

# Run it
print_tree()
```


